Utilization of Reductions and Abstraction Elimination in Typed Genetic Programming

### 2014-01-30 tex/gecco14/gecco14.tex
gecco finiš-him!

\newcommand{\Lets}{Let us\xspace}
\newcommand{\lets}{let us\xspace}
\newcommand{\lterm}{$\lambda$-term\xspace}
\newcommand{\lterms}{$\lambda$-terms\xspace}
\newcommand{\lhead}{$\lambda$-head\xspace}
\newcommand{\lheads}{$\lambda$-heads\xspace}
\newcommand{\la}{\leftarrow\xspace}
\newcommand{\Lp}  {\Lambda^{\prime}\xspace}
\newcommand{\tur}[3]{#1\vdash{}#2 \colon #3}
\newcommand{\turst}[3]{$#1\vdash{}#2:#3$\xspace}
\newcommand{\GMS}{\turst{\Gamma}{M}{\sigma}}
\newcommand{\atTree}{@-tree\xspace}
\newcommand{\setDots}[2]{ \lbrace #1 , \dots , #2 \rbrace}
\newcommand{\lh}[1]{\lambda #1}
\newcommand{\sexprTree}{sexpr-tree\xspace}
\newcommand{\SexprTree}{Sexpr-tree\xspace}
\newcommand{\then}{\Rightarrow\xspace}
\newcommand{\lamb}[2]{( \lambda \, #1 \, . \, #2 )}
\newcommand{\lam}[2]{\lambda \, #1 \, . \, #2}
\newcommand{\ST}{\mathop{\mathrm{ST}}}
\newcommand{\FV}{\mathop{\mathrm{FV}}}
\newcommand{\Scomb }{\mathbf{S}}
\newcommand{\Kcomb }{\mathbf{K}}
\newcommand{\Icomb }{\mathbf{I}}
\newcommand{\bbarr}{\twoheadrightarrow_\beta}
\newcommand{\barr}{\rightarrow_\beta}
\newcommand{\beq}{=_\beta}
\newcommand{\eearr}{\twoheadrightarrow_\eta}
\newcommand{\earr}{\rightarrow_\eta}
\newcommand{\eeq}{=_\eta}
\newcommand{\bearr}{\rightarrow_{\beta\eta}}
\newcommand{\bbeearr}{\twoheadrightarrow_{\beta\eta}}
\newcommand{\beeq}{=_{\beta\eta}}
\newcommand{\etar}{\twoheadrightarrow_\eta}
\newcommand{\ered}{$\eta$-reduction\xspace}
\newcommand{\bnf}{$\beta$-\textit{nf}\xspace}
\newcommand{\enf}{$\eta$-\textit{nf}\xspace}
\newcommand{\eenf}{$\eta^{-1}$-\textit{nf}\xspace}
\newcommand{\beenf}{$\beta\eta^{-1}$-\textit{nf}\xspace}
\newcommand{\benf}{$\beta\eta$-\textit{nf}\xspace}
\newcommand{\bredex}{$\beta$-redex\xspace} 
\newcommand{\lnf}{\textit{lnf}\xspace}
\newcommand{\Ae}{\mathop{\mathrm{\AE}}}
\newcommand{\Bcomb }{\mathbf{B}}   
\newcommand{\BBcomb }{\mathbf{B*}}
\newcommand{\Ccomb }{\mathbf{C}}   
\newcommand{\CCcomb }{\mathbf{C'}}
\newcommand{\SScomb }{\mathbf{S'}}
\newcommand{\ar}{\rightarrow\xspace}
\newcommand{\T}{\mathbb{T}\xspace}
\newcommand{\C}{\mathbb{C}\xspace}
\newcommand{\Real}{\mathbb{R}}
\newcommand{\gar}{\longmapsto}

@@

\numberofauthors{2}

\author{
% 1st. author
\alignauthor
Tom\'{a}\v{s} K\v{r}en\\
  \affaddr{Faculty of Mathematics and Physics}\\
  \affaddr{Charles University in Prague}\\
  \affaddr{Malostransk\'e n\'am\v{e}st\'\i~25, 11000,}\\
  \affaddr{Prague, Czech Republic}\\
  \email{tomkren@gmail.com}
% 2nd. author
\alignauthor
Roman Neruda\\
  \affaddr{Institute of Computer Science}\\
  \affaddr{Academy of Sciences of the Czech Republic}\\
  \affaddr{Pod Vod\'arenskou v\v{e}\v{z}\'\i~2, 18207,}\\
  \affaddr{Prague, Czech Republic}\\
  \email{roman@cs.cas.cz}
}

@@

\maketitle
\begin{abstract}
Lambda calculus representation of programs offers a more expressive alternative to traditional S-expressions. In this paper we discuss advantages of this representation coming from use of reductions (beta and eta) and how to overcome disadvantages caused by variables occurring in the programs by use of the abstraction elimination algorithm. We discuss the role of those reductions in the process of generating initial population and compare several crossover approaches including novel approach to crossover operator based both on reductions and abstraction elimination. The design goal of this operator is to turn the disadvantage of abstraction elimination  - possibly quadratic increase of program size - into a virtue; our approach leads to more crossover points. At the same time, utilization of reductions provides offspring of small sizes.
\end{abstract}

% A category with the (minimum) three required fields
\category{H.4}{\red{TODO}}{Miscellaneous}
%A category including the fourth, optional field follows...
\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]

\terms{\red{TODO}}

\keywords{\red{TODO, ToDo, todo}}

@@

Genetic programming (GP) represents an efficient method for automatic generating of programs by means of evolutionary techniques~\cite{koza92,koza03}. Early attempts to enhance the GP approach with the concept of types include the seminal work~\cite{montana95} where the ideas from Ada programming language were used to define a so-called strongly typed GP.   
Use of types naturally opens door to enriching S-expressions,
the traditional GP representation of individuals, with concepts from
lambda calculus, which is simple yet powerful functional, mathematical and programming language extensively used in type theory. Such attempts has shown to be successful \cite{yu01}.

\red{TODO....}

@@

\red{ve shrnutí paperu říct, že jeho tělo je tvořeno dvěma částma,
první se zabejvá generovánim a jeho vztahem k redukcím, druhá řeší křížení
s velkym důrazem na vyřešení problémů s lokálníma proměnejma.}

The rest of the paper is organized as follows: The next section briefly discusses related work in the field of typed GP, while section~\ref{preliminaries} introduces necessary notions. Main original results about search strategies in individual generating are described in section~\ref{approach}. Section~\ref{experiments} presents results of our method on three well-known tasks, and the paper is concluded by section~\ref{conclusions}.

\subsection{\red{Randombits}}
\red{

- Možná někam poznamenat že přínos toho článku je taky v tom že zpřístupňuje pojmy z teorie typovaného lc komunitě typovaného GP..

}

@@

\red{TODO zatim jen zkopčený z minula a dodaná diskuze 
zatim česky porovnávající nás a kombinatoři , 
opravit podle novýho kontextu}

Yu presents a GP system utilizing
polymorphic higher-order functions\footnote{Higher-order 
function is a function taking another function as 
input parameter.} and lambda abstractions  \cite{yu01}.
Important point of interest in this work is use of
\texttt{foldr} function as a tool for \textit{implicit recursion},
i.e. recursion without explicit recursive calls. 
The terminal set for constructing lambda abstraction subtrees 
is limited to use only constants and variables of that particular
lambda abstraction, i.e., outer variables are not allowed to be used
as terminals in this work. This is significant difference from our approach 
since we permit all well-typed normalized \lterms. From this difference also
comes different crossover operation. We focus more on term generating process; 
their term generation is performed in a similar way as the standard one, 
whereas our term generation also tries to utilize techniques of systematic enumeration.

Briggs and O’Neill present technique 
utilizing typed GP with combinators \cite{kes}.
The difference between approach presented in this work
and our approach is that in this work terms are generated
straight from \textit{library} of combinators and no lambda abstractions
are used. They are using more general polymorphic type system than us
-- the Hindley–Milner type system. They also discuss the 
properties of exhaustive enumeration of terms and compare it with GP search.  
They also present interesting concept of \textit{Generalized
genetic operator} based on term generation.

\red{

POZNAMKY PRO DISKUSI POROVNAVAJICI NÁS A BRIGGSE:

%Reprezentace v klasickym kozovy je S-expression.

The individual representation used in standard GP is S-expression. 
Basically there are two general approaches of generalizing S-expressions
into trees of better expressive power suitable for typed GP with higher 
order functions: We can enrich the method with the polymorphic combinators \cite{kes}
or we can enrich the method with the lambda abstractions and local variables
\cite{yu01}.

Pokud chceme dělat GP nad stromama větší vyjadřovací síly než maj klasický S-výrazy, tak máme v záse dvě přirozený možnosti pro obecný řešení: buď to dělat celý v (polymorfních) kombinátorech od začátku (a vyhnout se tak proměnejm a lambda abstrakcím), nebo to skusit v lambda kalkulu a nejak se vypořádat s proměnejma a lambdama.

Zdá se že použití kombinátorů má řadu výhod (vymenovat nějaký co jmenuje briggs), 
ale konzistentní prozkoumání obecnýho řešení lambdovýho charakteru určitě stojí 
taky za zkoužku, což podporují nasledující argumenty:

  (a) motivace pro práci s lambda termama je, že za nima košatá teorie, 
      která například popisuje redukce lambda termů (ty zajišťujou jak 
      zmenšení samotných stromu, tak to, že prohledávací prostor se zmenší, 
      díky tomu, že se různé stromy redukují na ten samej).
      (ALT znění: 
      pohovořit o tom, že normální formy zmenšujou prohledávaný prostor, 
      protže stačí brát v potaz třídy ekvivalence a ne zamotný termy 
      - reprezentanti těchto tříd ekvivalence jsou pak právě normální formy. )
      (ALT2: 
      Dost podstatný ospravedlnění lambda reprezentace je fakt, 
      že redukce (beta, eta) šlapou právě jen při tom, když máme lambdy.)

  (b) úspornost - abstrakce eliminací produkuje 
      termy až kvadraticky větší než je původní 
      term. Dá se očekávat, že i při přímém generování je použití kombinátorů 
      míň úsporný (proměná se může bez velký námahy objevit v hluboko vnořeném
      podtermu velmi jednoduše, zatímco pomocí kombinátorů se její hodnota musí do
      takto vnořeného podtermu složitě dopravit).

  (c) použití kombů vyžaduje silnější typovej sytém - 
      kombinátory musej bejt polymorfní a přidáváme je do stavební sady,
      lambdy v pohodě běžej na simply typed lc.

  (d) použití proměnných je typické pro lidi, proto by se mohlo sát že
      i programy generovaný systémem kterej s proměnejma počitá budou o něco 
      "lidštější"

Since abstr-elim a redukce neinvolvujou moc typově rilejtlý problémy, je ok celou věc demostrovat na jednoduchém typovém systému jakám je simply typed lc. Pro more soffisticated type systems would be situation similar.

Hlavním problémem lambda reprezentac proti kombinátorový je 
jak se vypořádat s problémy zpusobenými použitím 
localních proměných (aka jak křížit).

}

Binard and Felty use even 
stronger type system (\textit{System F}) \cite{binard2008genetic}.  
But with increasing power of the type system comes increasing difficulty of term generation.
For this reason evolution in this work takes interesting and nonstandard shape 
(fitness is associated with \textit{genes} which are evolved together with \textit{species}
which together participate in creation of individuals).
This differs from our approach, which tries to be generalization of
the standard GP\cite{koza92}.

In contrast with above mentioned works our approach uses very simple type system 
(simply typed lambda calculus) and concentrates on process of generation  
able to generate all possible well-typed normalized lambda terms. In order to do
so we use technique based on \textit{inhabitation machines} 
described by Barendregt \cite{barendregt10}.

@@

\red{TODO: ještě určitě projít a vylepšit; někde zestručnit někde rozšířit}

@@

\textit{Function application} and 
\textit{$\lambda$-abstraction} are concepts
well known from common programming languages. 
For example in JavaScript 
$(M~N)$ translates to expression \texttt{$M$($N$)} and
$\lamb{x}{M}$ translates to expression \texttt{function($x$)\{return $M$;\}}.
In other words, the function application 
corresponds to the act of supplying a function 
with an argument, and
the $\lambda$-abstraction is equivalent to 
\textit{anonymous function}\footnote{Apart from JavaScript, anonymous functions are common e.g. in Python and Ruby, 
they were recently introduced to C++, and they are expected to be supported in Java 8.}.

For better readability, 
$M_1~M_2~M_3~\dots~M_n$ is an abbreviation for
$(\dots((M_1~M_2)~M_3)~\dots~M_n)$
and $\lam{x_1 x_2 \dots x_n }{M}$ for 
$\lamb{x_1}{\lamb{x_2}{\dots\lamb{x_n}{M}\dots}}$.

\subsection{$\beta$-reduction}

In order to perform computation there must be some
mechanism for term evaluation. In $\lambda$-calculus there
is $\beta$-reduction for this reason.\\

\newcommand{\bRedex}{$\beta$-redex\xspace}
\newcommand{\bRedexes}{$\beta$-redexes\xspace}
\newcommand{\bArrow}{\rightarrow_\beta\xspace}
\newcommand{\eArrow}{\rightarrow_\eta\xspace}
\newcommand{\eeArrow}{\rightarrow_{\eta^{-1}}\xspace}

@@

\subsection{$\eta$-reduction}

Similarly as for $\beta$-reduction we can define $\eta$-reduction 
except that instead of \ref{eq:bRed} we use:

$$\lamb{x}{(M~x)} \eArrow M \textbf{ ~~~~if } x \not\in FV(M) $$~

@@

Relation $\bearr\;=\;\barr \cup \earr$. 
(Relation $R = \{\;(a,b)\;|\;a\;R\;b\;\}$.)

Similarly as for $\bbarr$ and $\beq$ we can define relations 
$\eearr$, $\eeq$,$\bbeearr$ and $\beeq$.

\subsection{$\eta^{-1}$-reduction}

$\eta^{-1}$-reduction (also called $\eta$-expansion) is 
the reduction converse to $\eta$-reduction.
Again it may be obtained by replacing \ref{eq:bRed}, now with:

$$M \eeArrow \lamb{x}{(M~x)} \textbf{ ~~~~if } x \not\in FV(M) $$

\subsection{Normal forms}

\red{asi řít tady v ty subsekci co že to je normalizace termu jednou větou}

~\begin{easylist}[enumerate]
& A \lterm is a \textit{$\beta$-normal form} (\bnf) 
if it does not have a $\beta$-redex as subterm.
& A \lterm M \textit{has} a \bnf if $M \beq N$
and $N$ is a \bnf.\\
\end{easylist}
A normal form may be thought of as a result of a term evaluation.

@@

\subsection{Types etc}

@@

Context is a basic type theoretic concept suitable as a typed alternative
for terminal and function set in standard GP. 
Notation $\Gamma,x:\sigma $ denotes $ \Gamma\cup\{(x:\sigma)\}$ 
such that $\Gamma$ does not contain any declaration with $x$ as subject.
We also write $x:\sigma \in \Gamma$ instead of $(x,\sigma) \in \Gamma$.

@@

Přeformulovat aby se to hodilo do tohodle kontextu....
Our goal in term generation is to produce terms $M$
for a given pair $\ul{\tau}{\Gamma}$
such that for each $M$ is $\tur{\Gamma}{M}{\tau}$.

@@

This is simple version of this process. More optimized version,
in the means of the size of resulting term and its performance
is following one, presented in \cite{jones87}.\\

As is stated in \cite{jones87},
the biggest disadvantage of this technique is that the translated
term is often much larger than in its lambda form --- the size of
the translated term can be proportional to the
square of the size of the original term.

But the advantage is also tempting --- no need to deal with variables
and lambda heads.

\red{tady je to v původnim kontextu, odebrat slova co se tam nehodí}

\section{Term generating and reductions}

@@

In \cite{barendregt10} is shown term generating grammar with 
following rules.

\red{předělat ten zápis, víc to okecat, říct že to je jakoby gramatika, 
formálně 2-level-gramatika do footnoty, pač má nekonečně neterminálu  
i terminálů. že neterminaly jsou dvojice (typ, kontext) a že terminaly jsou
symboly kterejma se zapisujou lambda termy}

\begin{align*}
( \alpha , \Gamma )  
&\gar
(~f~( \rho_1 , \Gamma )~\dots~( \rho_m , \Gamma )~)
\\& \textbf{if } \alpha \in A,
(f : \rho_1 \ar \dots \ar \rho_m \ar \alpha) \in \Gamma
\\ 
( \sigma \rightarrow \tau , \Gamma )  
&\gar
(~\lambda~x~.~( \tau ; \Gamma,x:\sigma )~)
&   
\end{align*}

The second rule can be replaced by more effective one.
$ 
( \tau_1 \ar \dots \ar \tau_n \ar \alpha , \Gamma )  
\gar \\
(~\lambda~x_1~\dots~x_n~.~
( \alpha ; \Gamma , x_1:\tau_1 , \dots , x_n:\tau_n  )~)
~~~~ \textbf{if } n > 0
$

This rule packs consecutive uses of the second rule into one use.
This is valid since the use of the second rule is deterministic;
it is used if and only if the non-terminal's type is not atomic.

Long normal form is a generalization of S-expressions into lambda calculus:
Let $\Gamma$ be a context satisfying the closure requirement, then only
the first rule will be applicable. 
$(\alpha, \Gamma)$ rewrites to $(~f~( \alpha, \Gamma )~\dots~( \alpha, \Gamma )~)$.
One can see that $f$ stands for 
the root node and that each $( \rho_i , \Gamma )$ will produce a direct subtree.

\red{
For context containing at least one higher order function we get special kind of root node containing lambda head with one direct subtree standing for body of 
the lambda function, v tomto tele se navíc mohou objevit proměnné 
definované v této hlavě (asi doplnit obrázkama).

Diskuze toho jakou přesnou strategií je nejvhodnější generovat
termy v lnf je beyond the scope of this paper, mužeme například použít
klasickou raped half-and half strategii. Footnotenout kraťoučkou
zmínku o gemetrický. 
}

@@

Every \lterm $M$ such that \GMS has its unique \lnf $L$, 
for which $L =_{\beta\eta} M$.
Therefore the computation performed by \lterm $M$ 
is not omitted, because it is the same computation
as the computation performed by \lterm $L$. \\

@@

For better clarity \lets show the \ered and \bredex using trees.\\
\ered: \\

\Tree [.$\lh{x}$ [.@ $M$ $x$ ] ] 
~~~~$\eArrow$~~~~
$M$

And \bredex:

\Tree [.@ [.$\lh{x}$ $M$ ] $N$ ] \\

\Lets assume that $P \eArrow Q$ creates a new \bredex $B$ in $Q$.

>>>>

\newcommand{\Lets}{Let us\xspace}
\newcommand{\lets}{let us\xspace}
\newcommand{\lterm}{$\lambda$-term\xspace}
\newcommand{\lterms}{$\lambda$-terms\xspace}
\newcommand{\lhead}{$\lambda$-head\xspace}
\newcommand{\lheads}{$\lambda$-heads\xspace}
\newcommand{\la}{\leftarrow\xspace}
\newcommand{\Lp}  {\Lambda^{\prime}\xspace}
\newcommand{\tur}[3]{#1\vdash{}#2 \colon #3}
\newcommand{\turst}[3]{$#1\vdash{}#2:#3$\xspace}
\newcommand{\GMS}{\turst{\Gamma}{M}{\sigma}}
\newcommand{\atTree}{@-tree\xspace}
\newcommand{\setDots}[2]{ \lbrace #1 , \dots , #2 \rbrace}
\newcommand{\lh}[1]{\lambda #1}
\newcommand{\sexprTree}{sexpr-tree\xspace}
\newcommand{\SexprTree}{Sexpr-tree\xspace}
\newcommand{\then}{\Rightarrow\xspace}
\newcommand{\lamb}[2]{( \lambda \, #1 \, . \, #2 )}
\newcommand{\lam}[2]{\lambda \, #1 \, . \, #2}
\newcommand{\ST}{\mathop{\mathrm{ST}}}
\newcommand{\FV}{\mathop{\mathrm{FV}}}
\newcommand{\Scomb }{\mathbf{S}}
\newcommand{\Kcomb }{\mathbf{K}}
\newcommand{\Icomb }{\mathbf{I}}
\newcommand{\bbarr}{\twoheadrightarrow_\beta}
\newcommand{\barr}{\rightarrow_\beta}
\newcommand{\beq}{=_\beta}
\newcommand{\eearr}{\twoheadrightarrow_\eta}
\newcommand{\earr}{\rightarrow_\eta}
\newcommand{\eeq}{=_\eta}
\newcommand{\bearr}{\rightarrow_{\beta\eta}}
\newcommand{\bbeearr}{\twoheadrightarrow_{\beta\eta}}
\newcommand{\beeq}{=_{\beta\eta}}
\newcommand{\etar}{\twoheadrightarrow_\eta}
\newcommand{\ered}{$\eta$-reduction\xspace}
\newcommand{\bnf}{$\beta$-\textit{nf}\xspace}
\newcommand{\enf}{$\eta$-\textit{nf}\xspace}
\newcommand{\eenf}{$\eta^{-1}$-\textit{nf}\xspace}
\newcommand{\beenf}{$\beta\eta^{-1}$-\textit{nf}\xspace}
\newcommand{\benf}{$\beta\eta$-\textit{nf}\xspace}
\newcommand{\bredex}{$\beta$-redex\xspace} 
\newcommand{\lnf}{\textit{lnf}\xspace}
\newcommand{\Ae}{\mathop{\mathrm{\AE}}}
\newcommand{\Bcomb }{\mathbf{B}}   
\newcommand{\BBcomb }{\mathbf{B*}}
\newcommand{\Ccomb }{\mathbf{C}}   
\newcommand{\CCcomb }{\mathbf{C'}}
\newcommand{\SScomb }{\mathbf{S'}}
\newcommand{\ar}{\rightarrow\xspace}
\newcommand{\T}{\mathbb{T}\xspace}
\newcommand{\C}{\mathbb{C}\xspace}
\newcommand{\Real}{\mathbb{R}}
\newcommand{\gar}{\longmapsto}
\newcommand{\bRedex}{$\beta$-redex\xspace}
\newcommand{\bRedexes}{$\beta$-redexes\xspace}
\newcommand{\bArrow}{\rightarrow_\beta\xspace}
\newcommand{\eArrow}{\rightarrow_\eta\xspace}
\newcommand{\eeArrow}{\rightarrow_{\eta^{-1}}\xspace}

@@

% \numberofauthors{2} 
%
%  \author{
%  % 1st. author
%  \alignauthor
%  Tom\'{a}\v{s} K\v{r}en\\
%    \affaddr{Faculty of Mathematics and Physics}\\
%    \affaddr{Charles University in Prague}\\
%    \affaddr{Malostransk\'e n\'am\v{e}st\'\i~25, 11000,}\\
%    \affaddr{Prague, Czech Republic}\\
%    \email{tomkren@gmail.com}
%  % 2nd. author
%  \alignauthor
%  Roman Neruda\\
%    \affaddr{Institute of Computer Science}\\
%    \affaddr{Academy of Sciences of the Czech Republic}\\
%    \affaddr{Pod Vod\'arenskou v\v{e}\v{z}\'\i~2, 18207,}\\
%    \affaddr{Prague, Czech Republic}\\
%    \email{roman@cs.cas.cz}
%  }

@@

\maketitle
\begin{abstract}
Lambda calculus representation of programs offers a more expressive alternative to traditional S-expressions. In this paper we discuss advantages of this representation coming from the use of reductions (beta and eta) and a way to overcome disadvantages caused by variables occurring in the programs by use of the abstraction elimination algorithm. We discuss the role of those reductions in the process of generating initial population and propose several crossover approaches, two of which are based on abstraction elimination and compared using the even parity benchmark problem. 
%including novel approach to crossover operator based both on reductions and abstraction elimination. 
%The design goal of this operator is to turn the disadvantage of abstraction elimination  - possibly quadratic increase of program size - into a virtue; our approach leads to more crossover points. At the same time, utilization of reductions provides offspring of small sizes.
\end{abstract}

%Starej abstrakt než se zjistilo že unpacking neni tak super:
%Lambda calculus representation of programs offers a more expressive alternative to traditional S-expressions. In this paper we discuss advantages of this representation coming from use of reductions (beta and eta) and how to overcome disadvantages caused by variables occurring in the programs by use of the abstraction elimination algorithm. We discuss the role of those reductions in the process of generating initial population and compare several crossover approaches including novel approach to crossover operator based both on reductions and abstraction elimination. The design goal of this operator is to turn the disadvantage of abstraction elimination  - possibly quadratic increase of program size - into a virtue; our approach leads to more crossover points. At the same time, utilization of reductions provides offspring of small sizes.

% A category with the (minimum) three required fields
\category{I.2.2}{Artificial Intelligence}{Automatic Programming}[Program synthesis]
%A category including the fourth, optional field follows...
\category{F.4.1}{Mathematical Logic and Formal Languages}{Mathematical Logic}[Lambda calculus and related systems]

\terms{Experimentation, Languages, Theory}

\keywords{genetic programming, lambda calculus}

@@

Genetic programming (GP) represents an efficient method for automatic generating of programs by means of evolutionary techniques~\cite{koza92,koza03}. Early attempts to enhance the GP approach with the concept of types include the seminal work~\cite{montana95} where the ideas from Ada programming language were used to define a so-called strongly typed GP.   
Usage of types naturally opens door to enriching S-expressions,
the traditional GP representation of individuals, with concepts from
lambda calculus, which is a simple yet powerful functional, mathematical and programming language extensively used in type theory. Such attempts have shown to be successful \cite{yu01}.

One of the motivations for using lambda calculus is that it is backed by a considerable theoretical background. From this background we can utilize several useful concepts such as reductions, normalization or abstraction elimination, and use them to enhance the standard GP algorithm. We use normalization to reduce the size of the search space of program trees and the abstraction elimination to prevent
the difficulties caused by the presence of variables in program trees.

In our system we use simply typed lambda calculus. Since abstraction elimination and reductions do not involve many type related issues it seems to us as a sufficient choice of type system. Techniques described here would be analogous for a more sophisticated type system and can be directly applied to such systems as the Hindley–Milner type system.

%\red{Možná někam poznamenat že přínos toho článku je taky v tom že zpřístupňuje pojmy z teorie typovaného lc komunitě typovaného GP esli to neni blbý..}

@@

The rest of the paper is organized as follows: The next section briefly discusses related work in the field of typed GP, while section~\ref{preliminaries} introduces necessary notions.
The body of the paper is presented in sections \ref{generating} and \ref{crossover}; the first discusses the term generating 
procedure and its relations
with reductions, the second describes several approaches to lambda term
crossover and how they solve the variable problem. Section~\ref{experiments} examines those ideas on the even parity benchmark problem, while the paper is concluded by section~\ref{conclusions}.

@@

Yu presents a GP system utilizing
polymorphic higher-order functions\footnote{Higher-order 
function is a function taking another function as 
input parameter.} and lambda abstractions  \cite{yu01}.
An important point of interest in the work is use of
\texttt{foldr} function as a tool for \textit{implicit recursion}, i.e. recursion without explicit recursive calls. The terminal set for constructing lambda abstraction subtrees  is limited to use only constants and variables of that particular lambda abstraction, i.e., outer variables are not allowed to be used as terminals in this work. This is a significant difference from our approach since we permit all well-typed normalized \lterms. This difference also causes the need for different crossover operators.

Briggs and O’Neill present a technique 
utilizing typed GP with combinators \cite{kes}.
The difference between approach presented in their work
and our approach is that they generate terms in a straightforward way directly from the library of combinators, without any use of lambda abstractions. They use the Hindley–Milner type system. They also present interesting concept of \textit{Generalized genetic operator} based on term generation. The combinator based genetic algorithm has the advantage of
avoiding the need of dealing with variables.

Binard and Felty use even stronger type system (\textit{System F}) \cite{binard2008genetic}. But with the increasing power of the type system also comes an increasing difficulty of term generation. For this reason, the evolution in this work takes interesting and nonstandard shape (fitness is associated with \textit{genes} which are evolved together with \textit{species} which together participate in creation of individuals). This differs from our approach, which tries to be a generalization of the standard GP\cite{koza92}.

In contrast with the above mentioned works, our approach uses very simple type system (simply typed lambda calculus) and concentrates on process of generation  
able to generate all possible well-typed normalized lambda terms. In order to do so we use technique based on \textit{inhabitation machines} described by Barendregt \cite{barendregt10}.

@@

@@

\textit{Function application} and 
\textit{$\lambda$-abstraction} are concepts
well known from common programming languages. 
For example, in JavaScript, 
$(M~N)$ translates to expression \texttt{$M$($N$)} and
$\lamb{x}{M}$ translates to expression \texttt{function($x$)\{return $M$;\}}.
In other words, the function application 
corresponds to the act of supplying a function 
with an argument, and
the $\lambda$-abstraction is equivalent to 
\textit{anonymous function}\footnote{Apart from JavaScript, anonymous functions are common e.g. in Python and Ruby, 
they were recently introduced to C++, and they are expected to be supported in Java 8.}.

For better readability, 
$M_1~M_2~M_3~\dots~M_n$ will be an abbreviation for
$(\dots((M_1~M_2)~M_3)~\dots~M_n)$,
and $\lam{x_1 x_2 \dots x_n }{M}$ will be an abbreviation for 
$\lamb{x_1}{\lamb{x_2}{\dots\lamb{x_n}{M}\dots}}$.

\subsection{Reductions}

In order to perform computation, there must be some
mechanism for term evaluation. In $\lambda$-calculus there
is a \mbox{$\beta$-reduction} procedure for this reason.\\

@@

Similarly as for $\beta$-reduction we can define $\eta$-reduction except that now it is defined as follows.  
$$\lamb{x}{(M~x)} \eArrow M \textbf{ ~~~~if } x \not\in FV(M) $$

@@

Relation $\bearr\;=\;\barr \cup \earr$. 
(Relation $R = \{\;(a,b)\;|\;a\;R\;b\;\}$.)
Similarly as for $\bbarr$ and $\beq$ we can define relations 
$\eearr$, $\eeq$,$\bbeearr$ and $\beeq$.\\

$\eta^{-1}$-reduction (also called $\eta$-expansion) is 
the reduction converse to $\eta$-reduction.
It is defined as follows.  
$$M \eeArrow \lamb{x}{(M~x)} \textbf{ ~~~~if } x \not\in FV(M) $$

\subsection{Normal forms}
\begin{definition}~

\begin{easylist}[enumerate]
& A \lterm is a \textit{$\beta$-normal form} (\bnf) 
if it does not have a $\beta$-redex as subterm.
& A \lterm M \textit{has} a \bnf if $M \beq N$
and $N$ is a \bnf.\\
\end{easylist}
\end{definition}
A normal form may be thought of as a result of a term evaluation. 
Normalization of the term $M$ is the process of finding normal form of $M$.

@@

\subsection{Types}

@@

Context is a basic type theoretic concept suitable as a typed alternative
for terminal, and function set in standard GP. 
Notation $\Gamma,x:\sigma $ denotes $ \Gamma\cup\{(x:\sigma)\}$ 
such that $\Gamma$ does not contain any declaration with $x$ as subject.
We also write $x:\sigma \in \Gamma$ instead of $(x,\sigma) \in \Gamma$.

@@

Our goal in term generation is to produce terms $M$
for a given pair $\ul{\tau}{\Gamma}$
such that for each $M$ is $\tur{\Gamma}{M}{\tau}$.

@@

As is stated in \cite{jones87},
the biggest disadvantage of this technique is that the translated
term is often much larger than in its lambda form --- the size of
the translated term can be proportional to the square of the size 
of the original term. But the advantage is also tempting --- no need to deal with variables and lambda abstractions.

The algorithm presented here is a simple version of this process. 
More optimized version (used in our \textit{hybrid} crossover described below), by means of the size of resulting term and its time performance is presented in \cite{jones87}.

\section{Term generating and reductions}
\label{generating}

@@

In \cite{barendregt10}, the following term generating 2-level grammar\footnote{2-level grammar (unlike a standard grammar) can have infinitely many non-terminals and its rule is better described as \textit{rule schema}. The original grammar described in \cite{barendregt10} has $( \sigma \rightarrow \tau , \Gamma ) \gar (~\lambda~x~.~( \tau ; \Gamma,x:\sigma )~)$ as second rule, but our grammar is equivalent -- it packs consecutive uses of this rule into our rule.} for generating all possible
terms in their \textit{long normal form} is described.

If we want to generate a term $M$ satisfying $\tur{\Gamma}{M}{\tau}$,
then the initial non-terminal is $(\tau, \Gamma)$.

For every $\alpha \in A$ and context $\Gamma$ and $f$ such that \\
$(f :$ $\rho_1 \ar \dots \ar \rho_m \ar \alpha$ $) \in \Gamma$ 
we have a rule of the first form:
$$
( \alpha , \Gamma ) \gar 
(~f~( \rho_1 , \Gamma )~\dots~( \rho_m , \Gamma )~)
$$

In other words, one possible way to construct a term of 
the atomic type $\alpha$ from the building symbols $\Gamma$
is to take a function symbol $f$ from $\Gamma$ with 
the return type $\alpha$ (or if $m=0$ a constant of the type $\alpha$),
put it into the root of the term and to generate 
a subtrees of appropriate types for each of $f$'s arguments
from the building symbols $\Gamma$.

For every atomic type $\alpha$, context $\Gamma$ and types
$\tau_1 , \dots , \tau_n$ (where $n > 0$) 
we have a rule of the second form:\\

$
( \tau_1 \ar \dots \ar \tau_n \ar \alpha , \Gamma )  
 \\ ~~~~~~~~\gar
(~\lambda~x_1~\dots~x_n~.~
( \alpha ; \Gamma , x_1:\tau_1 , \dots , x_n:\tau_n  )~)
$~\\

In other words, the way to construct a term of 
the function type $\tau_1 \ar \dots \ar \tau_n \ar \alpha$ 
from the building symbols $\Gamma$
is to create an anonymous function
with fresh variables $x_1, \dots , x_n$ in its head 
whose body is constructed as a term that has type $\alpha$
and is constructed using symbols from $\Gamma$ 
enriched with those new variables $x_1, \dots , x_n$,
which are associated with appropriate types.

% TODO dyštak zmínit že sme druhý pravidlo nahradili fikanou zkratkou
%The second rule can be replaced by more effective one.
%This rule packs consecutive uses of the second rule into one use. This is valid since the use of the second rule is deterministic; it is used if and only if the non-terminal's type is not atomic.

Long normal form is a generalization of S-expressions into lambda calculus:
Let $\Gamma$ be a context satisfying the closure requirement, then only
the first rule will be applicable. 
$(\alpha, \Gamma)$ rewrites to $(~f~( \alpha, \Gamma )~\dots~( \alpha, \Gamma )~)$.
One can see that $f$ stands for 
the root node and that each $( \alpha , \Gamma )$ will produce a direct subtree.

For a context containing at least one higher order function 
(or if the desired type of generated terms is a function type) we get special kind of root node containing lambda head with one direct subtree standing for body of the lambda function. Local variables defined in the lambda head may also appear in this body.

Discussion of what is the best strategy for browsing the search space given by above described grammar in order to generate \lnf terms is beyond the scope of this paper. Standard \textit{ramped half-and-half} strategy can be used for this purpose, or some other, 
which we discuss in \cite{nasecec}. In our experiments we use the \textit{geometric} strategy proposed in 
\cite{nasecec}.

@@

Every \lterm $M$ such that \GMS has its unique \lnf $L$, 
for which $L =_{\beta\eta} M$.
Therefore, the computation performed by \lterm $M$ 
is not omitted, because it is the same computation
as the computation performed by \lterm $L$. \\

@@

For better clarity \lets show the \ered and \bredex using \atTree{}s 
(a \lterm representation derived from the definition of \lterms 
where @ stands for \textit{function application}).\\
\textit{\ered}: \Tree [.$\lh{x}$ [.@ $M$ $x$ ] ] 
$\eArrow$
$M$ 
~~~~~~~~~~~
\textit{\bredex}: \Tree [.@ [.$\lh{x}$ $M$ ] $N$ ] \\
